Introducing ICBe: Very High Recall and Precision Event Extraction from Narratives about International Crises

### 2022-09-29 replication_paper/political_analysis_draft/ICBe_political_analysis_submission_rmd.tex
spell-check

\begin{abstract}
How do international crises unfold? We conceptualize of international
relations as a strategic chess game between adversaries and develop a
systematic way to measure pieces, moves, and gambits accurately and
consistently over a hundred years of history. We introduce a new
ontology and dataset of international events called ICBe based on a very
high-quality corpus of narratives from the International Crisis Behavior
(ICB) Project. We demonstrate that ICBe has higher coverage, recall, and
precision than existing state of the art datasets and conduct two
detailed case studies of the Cuban Missile Crisis (1962) and
Crimea-Donbas Crisis (2014). We further introduce two new event
visualizations (event icongraphy and crisis maps), an automated
benchmark for measuring event recall using natural language processing
(sythnetic narratives), and an ontology reconstruction task for
objectively measuring event precision. We make the data, online
appendix, replication material, and visualizations of every historical
episode available at a companion website www.crisisevents.org.
\end{abstract}

@@

If we could record every important interaction between countries in all
of diplomacy, military conflict, and international political economy,
how much unique information would this chronicle amount to, and how
surprised would we be to see something new? In other words, what is the
entropy of international relations? This record could in principle be
unbounded, but the central conceit of social science is that there are
structural regularities that limit what actors can do, their best
options, and even which actors are likely to survive (Brecher 1999;
Reiter 2015). If so, then these events can be systematically measured,
and accordingly, massive effort is expended in social science attempting
to record these regularities.\footnote{See work on crises (Brecher and
  Wilkenfeld 1982; Beardsley et al. 2020), militarized disputes (Palmer
  et al. 2021; Gibler 2018; Maoz et al. 2019), wars (Sarkees and Wayman
  2010; Reiter, Stam, and Horowitz 2016), organized violence (Ralph
  Sundberg and Mihai Croicu 2016; Pettersson and Eck 2018), political
  violence (Raleigh et al. 2010), sanctions (Felbermayr et al. 2020),
  trade (Barari and Kim, n.d.), and international agreements (Kinne
  2020; Owsiak, Cuttner, and Buck 2018; Vabulas and Snidal 2021),
  dispute resolution (Vabulas and Snidal 2021; Frederick, Hensel, and
  Macaulay 2017), and diplomacy (Moyer, Turner, and Meisel 2020; Sechser
  2011).} Thanks to improvements in natural language processing, more
open-ended efforts have begun to capture entire unstructured streams of
events from international news reports.\footnote{See Li et al. (2021);
  Halterman (2020); Brandt et al. (2018); Boschee et al. (2015); Hegre
  et al. (2020); Grant et al. (2017). On event-extraction from images
  and social-media see Zhang and Pan (2019) and Steinert-Threlkeld
  (2019).} How close these efforts are to accurately measuring all or
even most of what is essential in international relations is an open
empirical question, one for which we provide new evidence here.

Our contribution is a high coverage ontology and event dataset for key
historical episodes in 20th and 21st-century international relations. We
develop a large, flexible ontology of international events with the help
of both human coders and natural language processing. We apply it
sentence-by-sentence to an unusually high-quality corpus of historical
narratives of international crises (Brecher 1999; Brecher, James, and
Wilkenfeld 2000; Wilkenfeld and Brecher 2000; James 2019; Iakhnis and
James 2019). The result is a new lower bound estimate of how much
actually happens between states during pivotal historical episodes. We
then develop several methods for objectively gauging how well these
event codings reconstruct the information contained in the original
narrative. We conclude by benchmarking our event codings against several
current state-of-the-art event data collection efforts. We find that
existing systems produce sequences of events that do not contain enough
information to reconstruct the underlying historical episode. The
underlying fine-grained variation in international affairs is
unrecognizable through the lens of current quantification efforts.

This is a measurement paper that makes the following argument --- there
is a real-world unobserved latent concept known as international
relations, we propose a method for systematically measuring it, we
successfully apply this method producing a new large scale set of
measurements, those measurements exhibit several desirable kinds of
internal and external validity, and those measurements out-perform other
existing approaches. The article organizes that argument into eight
sections: task definition; corpus; priors/existing state of the art;
ICBe coding process; internal consistency; case study selection; recall;
and precision. A final section concludes.

@@

We consider the measurement task of abstracting discrete events about a
historical episode in international relations. The easiest way to convey
the task is with an example. Figure 1 shows a narrative account of the
Cuban Missile Crisis (1962) alongside a mapping from each natural
language sentence to discrete machine readable abstractive events.
Formally, a historical episode, \(H\), is demarcated by a period of time
\([T_{start}, T_{end}] \in T\), a set of Players \(p \in P\), and a set
of behaviors they undertook during that time \(b \in B\). International
Relations, \(IR\), is the system of regularities that govern the
strategic interactions that world actors make during a historical
episode, given their available options, preferences, beliefs, and
expectations of choices made by others. We observe neither \(H\) nor
\(IR\) directly. Rather the Historical Record, \(HR\), produces
documents \(d \in D\) containing some relevant and true (as well as
irrelevant and untrue) information about behaviors that were undertaken
recorded in the form of unstructured natural language text. The task is
to combine informative priors about \(IR\) with an unstructured corpus
\(D\) to produce a series of structured discrete events, \(e \in E\),
that have high coverage, precision, and recall over what actually took
place in history, \(H\).

@@

We base our informed priors primarily on two sources of information. The
first is the extensive existing measurement efforts of IR which we
provide citations to alongside each concept. Second, we performed
preliminary natural language processing of the corpus and identified
named entities and behaviors mentioned in the text. Verbs were matched
to the most likely definition found in Wordnet (Miller 1995), tallied,
and then aggregated into a smaller number hypernyms balancing conceptual
detail and manageable sparsity for human coding (SI Appendix 1.2).

The second axis (columns) compares the very high ontological coverage of
ICBe to existing state of the art systems in production and with global
coverage. They begin with our contribution ICBe, alongside other
event-level datasets including CAMEO dictionary lookup based systems
(Historical Phoenix (Althaus et al. 2019); ICEWS (Boschee et al. 2015;
Hegre et al. 2020); Terrier (Grant et al. 2017)), the Militarized
Interstate Disputes Incidents dataset, and the UCDP-GED dataset (Ralph
Sundberg and Mihai Croicu 2016; Pettersson and Eck 2018; Sundberg and
Melander 2013).\footnote{Additional relevant but dated or too small of
  an overlap in domain include BCOW (Leng and Singer 1988), WEIS
  (McClelland 1978), CREON (Hermann 1984), CASCON (Bloomfield and
  Moulton 1989), SHERFACS (Sherman 2000), Real-Time Phoenix (Brandt et
  al. 2018), and COfEE (Balali, Asadpour, and Jafari 2021) (see
  histories in Merritt (1994) and Schrodt and Hall (2006)).} The final
set of columns compares episode-level datasets beginning with the
original ICB project (Brecher et al., n.d.; Brecher and Wilkenfeld 1982;
Beardsley et al. 2020); the Militarized Interstate Disputes dataset
(Palmer et al. 2021; Gibler 2018; Braithwaite 2010, 2009), and the
Correlates of War (Sarkees and Wayman 2010). With the exception of large
scale CAMEO dictionary based systems, the existing state of the art
quantitative datasets ignore the vast majority of the information
content found in international relations.\footnote{See (Balali,
  Asadpour, and Jafari 2021) for a recent review of ontological depth
  and availability of Gold Standard example text.}

@@

The ICBe ontology follows a hierarchical design philosophy where a
smaller number of significant decisions are made early on and then
progressively refined into more specific details (Brust and Denzler
2020).\footnote{This process quickly focuses the coder on a smaller
  number of relevant options while also allowing them to apply multiple
  tags if the sentence explicitly includes more than one or there is
  insufficient evidence to choose only one tag. The guided coding
  process also allows for the possibility that earlier coarse decisions
  have less error than later fine-grained decisions.} Each coder was
instructed to first thoroughly read the full crisis narrative and then
presented with a custom graphical user interface (SI Appendix 2.1).
Coders then proceeded sentence by sentence, choosing the number of
events (0-3) that occurred, the highest behavior (thought, speech, or
activity), a set of players (\(P\)), whether the means were primarily
armed or unarmed, whether there was an increase or decrease in
aggression (uncooperative/escalating or cooperative/de-escalating), and
finally one or more non-mutually exclusive specific activities. Some
additional details like location and timing information was always
collected while other details were only collected if appropriate,
e.g.~force size, fatalities, domains, units, etc. A unique feature of
the ontology is that thought, speech, and do behaviors can be nested
into combinations, e.g.~an offer for the U.S.S.R. to remove missiles
from Cuba in exchange for the U.S. removing missiles from Turkey.
Through compounding, the ontology can capture what players were said to
have known, learned, or said about other specific fully described
actions.

@@

We attribute the remaining disagreement primarily to three sources.
First, we required coders to rate and justify their confidence in the
coding. They reported low confidence for 20\% of sentences; 45\% of
those were due to a mismatch between the ontology and the text (``survey
doesn't fit event'') and 46\% were from a lack of information or
confused writing in the source text (40\% ``more knowledge needed'', 6\%
``confusing sentence''). Observed disagreement varied predictably with
self reported confidence (SI Appendix 2.4). Second, as intended,
agreement is higher (75-80\%) for questions with fewer options near the
root of the ontology compared to agreement for questions near the leafs
of the ontology (50\%-60\%). Third, individual coders exhibiting
nontrivial coding styles, e.g.~some more expressive applying many tags
per concept while others focused on only the single best match. We
further observed unintended synonymity, e.g.~the same information can be
framed as either a threat to do something or a promise not to do
something.

@@

Recall measures the share of desired information recovered by a sequence
of coded events, \(Pr(E|H)\), and is poorly defined for historical
episodes. First, there is no genuine ground truth about what occurred,
only surviving texts about it. Second, there is no \textit{a priori}
guide to what information is necessary detail and what is ignorable
trivia. History suffers from what is known as the Coastline Paradox
(Mandelbrot 1983) --- it has a fractal dimension greater than one such
that the more you zoom in the more detail you will find about individual
events and in between every two discrete events. The ICBe ontology is a
proposal about what information is important, but we need an independent
benchmark to evaluate whether that proposal is a good one that allows
for comparing proposals from event projects that had different goals. We
need a yardstick for history.

Our strategy for dealing with both problems is a plausibly objective
yardstick called a synthetic historical narrative. For both case
studies, we collect a large diverse corpus of narratives spanning
timelines, encyclopedia entries, journal articles, news reports,
websites, and government documents. Using natural language processing
(fully described in SI Appendix 3.1), we identify details that appear
across multiple accounts. The more accounts that mention a detail, the
more central it is to understanding the true historical episode. The
theoretical motivation is that authors face word limits which force them
to pick and choose which details to include, and they choose details
which serve the specific context of the document they are producing.
With a sufficiently large and diverse corpus of documents, we can vary
the context while holding the overall episode constant and see which
details tend to be invariant to context. Intuitively, a high quality
event dataset should have high recall for context invariant details both
because of their broader relevance and also because they are easier to
find in source material.

Synthetic historical narratives for Cuban Missiles (51 events drawn from
2020 documents) and Crimea-Donbas (30 events drawn from 971 documents)
appear in Figure 2. Each row represents a detail which appeared in at
least five documents along with an approximate start date, a hand
written summary, the number of documents it was mentioned in, and
whether it could be identified in the text of our ICB corpus, in our
ICBe events, or any of the competing systems.

From them, we draw several stylized facts. First, there is substantial
variation in which details any one document will choose to include. Our
ground truth ICB narratives included 17/51 and 23/30 of the events from
the synthetic narrative, while including other details that are not in
the synthetic narrative. Second, mentions of a detail across accounts is
exponentially distributed with context invariant details appearing
dozens to hundreds of times more than context dependent details. Third,
crisis start and stop dates are arbitrary and the historical record
points to many precursor events as necessary detail for understanding
later events, e.g.~the U.S. was in a \textit{de facto} grey scale war
with Cuba before it invited Soviet military protection (Cormac and
Aldrich 2018) and Ukraine provided several security guarantees to Russia
that were potentially undone, e.g.~a long term lease on naval facilities
in Crimea. Fourth, we find variation between the two cases. Cuban
Missiles has a cleaner canonical end with the Soviets agreeing to
withdraw missiles while Crimea-Donbas meekly ends with a second cease
fire agreement (Minsk II) but continued fighting. The canonical
narrative of Cuban Missile also includes high level previously
classified details, while the more recent Crimea-Donbas case reflects
primarily public reporting.

We find substantive variation in recall across systems. Recall for each
increases in the number of document mentions which is an important sign
of validity for both them and our benchmark. The one outlier is Phoenix
which is so noisy that it's flat to decreasing in mentions. The two
episode level datasets have very low coverage of contextual details. The
two other dictionary systems ICEWs and Terrier have high coverage, with
ICEWs outperforming Terrier. ICBe strictly dominates all of the systems
but ICEWs in recall though we note that the small sample sizes mean
these systems should be considered statistically indistinguishable.
Importantly our corpus of ICB narratives has very high recall of
frequently mentioned details giving us confidence in how those summaries
were constructed, and ICBe lags only slightly behind showing that it
left very little additional information on the table.

@@

The other side of event measurement is precision, the degree to which a
sequence of events correctly and usefully describes the information in
history, \(Pr(H|E)\). It does little good to recall a historical event
but too vaguely (e.g.~MIDs describes the Cuban Missile crisis as a
blockade, a show of force, and a stalemate) or with too much error
(e.g.~ICEWS records 263 ``Detonate Nuclear Weapons'' events between
1995-2019) to be useful for downstream applications. ICBe's ontology and
coding system is designed to strike a balance so that the most important
information is recovered accurately but also abstracted to a level that
is still useful and interpretable. You should be able to lay out events
of a crisis on a timeline, as in Figure 3, and read off the macro
structure of an episode from each individual move. We call this
visualization a crisis map, a directed graph intersected with a
timeline, and provide crisis maps for every event dataset for each case
study in SI Appendix 4.3 and 4.4 and for all crises on the companion
website.

We further want to verify individual event codings, which we can do in
the case of ICBe because each event is mapped to a specific span of
text. We develop the iconography system for presenting event codings as
coherent statements that can be compared side by side to the original
source narrative as for Cuban Missiles (Figure 1), Crimea-Donbas (SI
Appendix 4.1), and for every case on the companion website. We further
provide a stratified sample of event codings alongside their source text
(SI Appendix 4.2).

We find both the visualizations of macro structure and head-to-head
comparisons of ICBe codings to the raw text to strongly support the
quality of ICBe, but as with recall we seek a more objective detached
universal benchmark. Our proposed measure is a reconstruction task to
see whether our intended ontology can be recovered through only
unsupervised clustering of sentences they were applied to. Figure 4
shows the location of every sentence from the ICBe corpus in semantic
space as embeded using the same large language model as before, and the
median location of each ICBe event tag applied to those
sentences.\footnote{We preprocess sentences to replace named entities
  with a generic Entity token.} Labels reflect the individual leaves of
the ontology and colors reflect the higher level coarse branch nodes of
the ontology. If ICBe has high precision, substantively similar tags
ought to have been applied to substantively similar source text, which
is what we see both in two dimensions in the main plot and via
hierarchical clustering on all dimensions in the dendrogram along the
righthand side.\footnote{Hierchcial clustering on cosine similarity and
  with Ward's method.}

Finally, how does ICBe's precision compare to the existing state of the
art? The crisis-maps reveal the episode level datasets like MIDs or the
original ICB are too sparse and vague to reconstruct the structure of
the crisis (SI Appendix 4.3 and 4.4). On the other end of the spectrum,
the high recall dictionary based event datasets like Terrier and ICEWs
produce so many noisy events (several hundreds thousands) that even with
heavy filtering their crisis maps are completely unintelligible.
Further, because of copyright issues, none of these datasets directly
provide the original text spans making event level precision difficult
to verify.

However, given their high recall on our task and the global and
real-time coverage of dictionary based event systems, we want to take
seriously the possibility that some functional transformation could
recover the precision of ICBe. For example, Terechshenko (2020) attempts
to correct for the mechanically increasing amount of news coverage each
year by detrending violent event counts from Phoenix using a human coded
baseline. Others have focused on verifying precision for ICEWs on
specific subsets of details against known ground truths,
e.g.~geolocation (Cook and Weidmann 2019), protest events (80\%) (Wüest
and Lorenzini 2020), anti-government protest networks (46.1\%) (Jäger
2018).

We take the same approach here in Figure 5, selecting four specific
CAMEO event codings and checking how often they reflect a true real
world event. We choose four event types around key moments in the
crisis. The start of the crisis revolves around Ukraine backing out of
trade deal with the EU in favor of Russia, but ``sign formal agreement''
events act more like a topic detector with dozens of events generated by
discussions of a possible agreement but not the actual agreement which
never materialized. The switch is caught by the ``reject plan, agreement
to settle dispute'', but also continues for Victor Yanukovych for even
after he was removed from power because of articles retroactively
discussing the cause of his of his removal. Events for ``use
conventional military force'' capture a threshold around the start of
hostilities and who the participants were but not any particular battles
or campaigns. Likewise, ``impose embargo, boycott, or sanctions''
captures the start of waves of sanctions and from who but are
effectively constantly as the news coverage does not distinguish between
subtle changes or additions. In sum, dictionary based methods on news
corpora tend to have high recall because they parse everything in the
news, but for the same reason their specificity for most event types is
too low to back out individual chess like sequencing that ICBe aims to
record.

@@

\begin{figure}[H]
\caption{ICBe event codings in comparison to Semantic Embeddings from source sentences\label{fig:semantic_embeddings}}
\centering{\includegraphics[width=17cm]{./p_semantic_embeddings_dendro.png}}
\textit{Notes: Dots represent individual ICB narrative sentences, as embeded by the Paraphrase-MPNET-base-v2 large language model and flattened into two dimensions with UMAP. Text labels reflect individual leaves of the ICBe ontology, and colors represent intermediate branches of the ontology. Label placement is the median of all of the sentences that tag was applied to by the coders. The dendrogram shows hiearchical clustering of the tags. If ICBe precision is high, then the sentences that tags were applied to ought to say similar things and the intended shape of the ontology ought to be visually recognizable. }
\end{figure}
\clearpage

>>>>

\begin{abstract}
How do international crises unfold? We conceptualize international
relations as a strategic chess game between adversaries and develop a
systematic way to measure pieces, moves, and gambits accurately and
consistently over a hundred years of history. We introduce a new
ontology and dataset of international events called ICBe based on a very
high-quality corpus of narratives from the International Crisis Behavior
(ICB) Project. We demonstrate that ICBe has higher coverage, recall, and
precision than existing state of the art datasets and conduct two
detailed case studies of the Cuban Missile Crisis (1962) and the
Crimea-Donbas Crisis (2014). We further introduce two new event
visualizations (event iconography and crisis maps), an automated
benchmark for measuring event recall using natural language processing
(synthetic narratives), and an ontology reconstruction task for
objectively measuring event precision. We make the data, supplementary
appendix, replication material, and visualizations of every historical
episode available at a companion website www.crisisevents.org.
\end{abstract}

@@

If we could record every important interaction between countries in all
of diplomacy, military conflict, and international political economy,
how much unique information would this chronicle amount to, and how
surprised would we be to see something new? In other words, what is the
entropy of international relations? This record could in principle be
unbounded, but the central conceit of social science is that there are
structural regularities that limit what actors can do, their best
options, and even which actors are likely to survive (Brecher 1999;
Reiter 2015). If so, then these events can be systematically measured,
and accordingly, massive effort is expended in social science attempting
to record these regularities.\footnote{See work on crises (Brecher and
  Wilkenfeld 1982; Beardsley et al. 2020), militarized disputes (Palmer
  et al. 2021; Gibler 2018; Maoz et al. 2019), wars (Sarkees and Wayman
  2010; Reiter, Stam, and Horowitz 2016), organized violence (Ralph
  Sundberg and Mihai Croicu 2016; Pettersson and Eck 2018), political
  violence (Raleigh et al. 2010), sanctions (Felbermayr et al. 2020),
  trade (Barari and Kim, n.d.), and international agreements (Kinne
  2020; Owsiak, Cuttner, and Buck 2018; Vabulas and Snidal 2021),
  dispute resolution (Vabulas and Snidal 2021; Frederick, Hensel, and
  Macaulay 2017), and diplomacy (Moyer, Turner, and Meisel 2020; Sechser
  2011).} Thanks to improvements in natural language processing, more
open-ended efforts have begun to capture entire unstructured streams of
events from international news reports.\footnote{See Li et al. (2021);
  Halterman (2020); Brandt et al. (2018); Boschee et al. (2015); Hegre
  et al. (2020); Grant et al. (2017). On event extraction from images
  and social media see Zhang and Pan (2019) and Steinert-Threlkeld
  (2019).} How close these efforts are to accurately measuring all or
even most of what is essential in international relations is an open
empirical question, one for which we provide new evidence here.

Our contribution is a high coverage ontology and event dataset for key
historical episodes in 20th and 21st-century international relations. We
develop a large, flexible ontology of international events with the help
of both human coders and natural language processing. We apply it
sentence-by-sentence to an unusually high-quality corpus of historical
narratives of international crises (Brecher 1999; Brecher, James, and
Wilkenfeld 2000; Wilkenfeld and Brecher 2000; James 2019; Iakhnis and
James 2019). The result is a new lower bound estimate of how much
actually happens between states during pivotal historical episodes. We
then develop several methods for objectively gauging how well these
event codings reconstruct the information contained in the original
narrative. We conclude by benchmarking our event codings against several
current state of the art event data collection efforts. We find that
existing systems produce sequences of events that do not contain enough
information to reconstruct the underlying historical episode. The
underlying fine-grained variation in international affairs is
unrecognizable through the lens of current quantification efforts.

This is a measurement paper that makes the following argument --- there
is a real-world unobserved latent concept known as international
relations, we propose a method for systematically measuring it, we
successfully apply this method producing a new large-scale set of
measurements, and those measurements exhibit several desirable kinds of
internal and external validity, and those measurements out-perform other
existing approaches. The article organizes that argument into eight
sections: task definition; corpus; priors/existing state of the art;
ICBe coding process; internal consistency; case study selection; recall;
and precision. A final section concludes.

@@

We consider the measurement task of absstracting discrete events about a
historical episode in international relations. The easiest way to convey
the task is with an example. Figure 1 shows a narrative account of the
Cuban Missile Crisis (1962) alongside a mapping from each natural
language sentence to discrete machine-readable abstractive events.
Formally, a historical episode, \(H\), is demarcated by a period of time
\([T_{start}, T_{end}] \in T\), a set of Players \(p \in P\), and a set
of behaviors they undertook during that time \(b \in B\). International
Relations, \(IR\), is the system of regularities that govern the
strategic interactions that world actors make during a historical
episode, given their available options, preferences, beliefs, and
expectations of choices made by others. We observe neither \(H\) nor
\(IR\) directly. Rather the Historical Record, \(HR\), produces
documents \(d \in D\) containing some relevant and true (as well as
irrelevant and untrue) information about behaviors that were undertaken
recorded in the form of unstructured natural language text. The task is
to combine informative priors about \(IR\) with an unstructured corpus
\(D\) to produce a series of structured discrete events, \(e \in E\),
that have high coverage, precision, and recall over what actually took
place in history, \(H\).

@@

We base our informed priors primarily on two sources of information. The
first is the extensive existing measurement efforts of IR which we
provide citations alongside each concept. Second, we performed
preliminary natural language processing of the corpus and identified
named entities and behaviors mentioned in the text. Verbs were matched
to the most likely definition found in Wordnet (Miller 1995), tallied,
and then aggregated into a smaller number of hypernyms balancing
conceptual detail and manageable sparsity for human coding (SI Appendix
1.2).

The second axis (columns) compares the very high ontological coverage of
ICBe to existing state of the art systems in production and with global
coverage. They begin with our contribution ICBe, alongside other
event-level datasets including CAMEO dictionary lookup-based systems
(Historical Phoenix (Althaus et al. 2019); ICEWS (Boschee et al. 2015;
Hegre et al. 2020); Terrier (Grant et al. 2017)), the Militarized
Interstate Disputes Incidents dataset, and the UCDP-GED dataset (Ralph
Sundberg and Mihai Croicu 2016; Pettersson and Eck 2018; Sundberg and
Melander 2013).\footnote{Other datasets that insufficiently overlap
  ICBe's domain for comparison include BCOW (Leng and Singer 1988), WEIS
  (McClelland 1978), CREON (Hermann 1984), CASCON (Bloomfield and
  Moulton 1989), SHERFACS (Sherman 2000), Real-Time Phoenix (Brandt et
  al. 2018), and COfEE (Balali, Asadpour, and Jafari 2021) (see
  histories in Merritt (1994) and Schrodt and Hall (2006)).} The final
set of columns compares episode-level datasets beginning with the
original ICB project (Brecher et al., n.d.; Brecher and Wilkenfeld 1982;
Beardsley et al. 2020); the Militarized Interstate Disputes dataset
(Palmer et al. 2021; Gibler 2018; Braithwaite 2010, 2009), and the
Correlates of War (Sarkees and Wayman 2010). With the exception of
large-scale CAMEO dictionary-based systems, the existing state of the
art quantitative datasets ignore the vast majority of the information
content found in international relations.\footnote{See (Balali,
  Asadpour, and Jafari 2021) for a recent review of ontological depth
  and availability of Gold Standard example text.}

@@

The ICBe ontology follows a hierarchical design philosophy where a
smaller number of significant decisions are made early on and then
progressively refined into more specific details (Brust and Denzler
2020).\footnote{This process quickly focuses the coder on a smaller
  number of relevant options while also allowing them to apply multiple
  tags if the sentence explicitly includes more than one or there is
  insufficient evidence to choose only one tag. The guided coding
  process also allows for the possibility that earlier coarse decisions
  have less error than later fine-grained decisions.} Each coder was
instructed to first thoroughly read the full crisis narrative and then
presented with a custom graphical user interface (SI Appendix 2.1).
Coders then proceeded sentence by sentence, choosing the number of
events (0-3) that occurred, the highest behavior (thought, speech, or
activity), a set of players (\(P\)), whether the means were primarily
armed or unarmed, whether there was an increase or decrease in
aggression (uncooperative/escalating or cooperative/de-escalating), and
finally one or more non-mutually exclusive specific activities. Some
additional details like location and timing information were always
collected while other details were only collected if appropriate,
e.g.~force size, fatalities, domains, units, etc. A unique feature of
the ontology is that thought, speech, and do behaviors can be nested
into combinations, e.g.~an offer for the U.S.S.R. to remove missiles
from Cuba in exchange for the U.S. removing missiles from Turkey.
Through compounding, the ontology can capture what players were said to
have known, learned, or said about other specific fully described
actions.

@@

We attribute the remaining disagreement primarily to three sources.
First, we required coders to rate and justify their confidence in the
coding. They reported low confidence for 20\% of sentences; 45\% of
those were due to a mismatch between the ontology and the text (``survey
doesn't fit event'') and 46\% were from a lack of information or
confused writing in the source text (40\% ``more knowledge needed'', 6\%
``confusing sentence''). Observed disagreement varied predictably with
self-reported confidence (SI Appendix 2.4). Second, as intended,
agreement is higher (75-80\%) for questions with fewer options near the
root of the ontology compared to agreement for questions near the leaves
of the ontology (50\%-60\%). Third, individual coders exhibit nontrivial
coding styles, e.g.~some more expressive coders applied many tags per
concept while others focused on only the single best match. We further
observed unintended synonymity, e.g.~the same information can be framed
as either a threat to do something or a promise not to do something.

@@

Recall measures the share of desired information recovered by a sequence
of coded events, \(Pr(E|H)\), and is poorly defined for historical
episodes. First, there is no genuine ground truth about what occurred,
only surviving texts about it. Second, there is no \textit{a priori}
guide to what information is necessary detail and what is ignorable
trivia. History suffers from what is known as the Coastline Paradox
(Mandelbrot 1983) --- it has a fractal dimension greater than one such
that the more you zoom in, the more detail you will find about
individual events and in between every two discrete events. The ICBe
ontology is a proposal about what information is important, but we need
an independent benchmark to evaluate whether that proposal is a good one
and that allows for comparing proposals from event projects that had
different goals. We need a yardstick for history.

Our strategy for dealing with both problems is a plausibly objective
yardstick called a synthetic historical narrative. For both case
studies, we collect a large diverse corpus of narratives spanning
timelines, encyclopedia entries, journal articles, news reports,
websites, and government documents. Using natural language processing
(fully described in SI Appendix 3.1), we identify details that appear
across multiple accounts. The more accounts that mention a detail, the
more central it is to understanding the true historical episode. The
theoretical motivation is that authors face word limits which force them
to pick and choose which details to include, and they choose details
that serve the specific context of the document they are producing. With
a sufficiently large and diverse corpus of documents, we can vary the
context while holding the overall episode constant and see which details
tend to be invariant to context. Intuitively, a high-quality event
dataset should have high recall for context invariant details both
because of their broader relevance and also because they are easier to
find in source material.

Synthetic historical narratives for Cuban Missiles (51 events drawn from
2020 documents) and Crimea-Donbas (30 events drawn from 971 documents)
appear in Figure 2. Each row represents a detail that appeared in at
least five documents along with an approximate start date, a handwritten
summary, the number of documents it was mentioned in, and whether it
could be identified in the text of our ICB corpus, in our ICBe events,
or any of the competing systems.

From them, we draw several stylized facts. First, there is substantial
variation in which details any one document will choose to include. Our
ground truth ICB narratives contain 17/51 and 23/30 of the events from
the synthetic narrative and include other details that are not in the
synthetic narrative. Second, mentions of a detail across accounts are
exponentially distributed with context-invariant details appearing
dozens to hundreds of times more than context-dependent details. Third,
crisis start and stop dates are arbitrary, and the historical record
points to many precursor events as necessary detail for understanding
later events, e.g.~the U.S. was in a \textit{de facto} grey-scale war
with Cuba before it invited Soviet military protection (Cormac and
Aldrich 2018) and Ukraine provided several security guarantees to Russia
that were potentially undone, e.g.~a long term lease on naval facilities
in Crimea. Fourth, we find variation between the two cases. Cuban
Missiles has a cleaner canonical end with the Soviets agreeing to
withdraw missiles while Crimea-Donbas meekly ends with a second
cease-fire agreement (Minsk II) but continued fighting. The canonical
narrative of Cuban Missiles also includes high-level previously
classified details, while the more recent Crimea-Donbas case reflects
primarily public reporting.

We find substantive variation in recall across systems. Recall for each
increases in the number of document mentions which is an important sign
of validity for both them and our benchmark. The one outlier is Phoenix
which is so noisy that it's flat to decreasing in mentions. The two
episode-level datasets have very low coverage of contextual details. The
two other dictionary systems ICEWs and Terrier have high coverage, with
ICEWs outperforming Terrier. ICBe strictly dominates all of the systems
but ICEWs in recall though we note that the small sample sizes mean
these systems should be considered statistically indistinguishable.
Importantly our corpus of ICB narratives has very high recall of
frequently mentioned details giving us confidence in how those summaries
were constructed, and ICBe lags only slightly behind showing that it
left very little additional information on the table.

@@

The other side of event measurement is precision, the degree to which a
sequence of events correctly and usefully describes the information in
history, \(Pr(H|E)\). It does little good to recall a historical event
but too vaguely (e.g.~MIDs describes the Cuban Missile Crisis as a
blockade, a show of force, and a stalemate) or with too much error
(e.g.~ICEWS records 263 ``Detonate Nuclear Weapons'' events between
1995-2019) to be useful for downstream applications. ICBe's ontology and
coding system are designed to strike a balance so that the most
important information is recovered accurately but also abstracted to a
level that is still useful and interpretable. You should be able to lay
out the events of a crisis on a timeline, as in Figure 3, and read off
the macrostructure of an episode from each individual move. We call this
visualization a crisis map, a directed graph intersected with a
timeline, and provide crisis maps for every event dataset for each case
study in SI Appendix 4.3 and 4.4 and for all crises on the companion
website.

We further want to verify individual event codings, which we can do in
the case of ICBe because each event is mapped to a specific span of
text. We develop the iconography system for presenting event codings as
coherent statements that can be compared side by side to the original
source narrative for Cuban Missiles (Figure 1), Crimea-Donbas (SI
Appendix 4.1), and for every case on the companion website. We further
provide a stratified sample of event codings alongside their source text
(SI Appendix 4.2).

We find both the visualizations of macrostructure and head-to-head
comparisons of ICBe codings to the raw text to strongly support the
quality of ICBe, but as with recall, we seek a more objective detached
universal benchmark. Our proposed measure is a reconstruction task to
see whether our intended ontology can be recovered through only
unsupervised clustering of sentences they were applied to. Figure 4
shows the location of every sentence from the ICBe corpus in semantic
space as embedded using the same large language model as before, and the
median location of each ICBe event tag applied to those
sentences.\footnote{We preprocess sentences to replace named entities
  with a generic Entity token.} Labels reflect the individual leaves of
the ontology and colors reflect the higher level coarse branch nodes of
the ontology. If ICBe has high precision, substantively similar tags
ought to have been applied to substantively similar source text, which
is what we see both in two dimensions in the main plot and via
hierarchical clustering on all dimensions in the dendrogram along the
righthand side.\footnote{Hierchcial clustering on cosine similarity and
  with Ward's method.}

Finally, how does ICBe's precision compare to the existing state of the
art? The crisis maps reveal the episode-level datasets like MIDs or the
original ICB are too sparse and vague to reconstruct the structure of
the crisis (SI Appendix 4.3 and 4.4). On the other end of the spectrum,
the high recall dictionary-based event datasets like Terrier and ICEWs
produce so many noisy events (several hundred thousand) that even with
heavy filtering their crisis maps are completely unintelligible.
Further, because of copyright issues, none of these datasets directly
provide the original text spans making event-level precision difficult
to verify.

However, given their high recall on our task and the global and
real-time coverage of dictionary-based event systems, we want to take
seriously the possibility that some functional transformation could
recover the precision of ICBe. For example, Terechshenko (2020) attempts
to correct for the mechanically increasing amount of news coverage each
year by detrending violent event counts from Phoenix using a human-coded
baseline. Others have focused on verifying precision for ICEWs on
specific subsets of details against known ground truths,
e.g.~geolocation (Cook and Weidmann 2019), protest events (80\%) (Wüest
and Lorenzini 2020), anti-government protest networks (46.1\%) (Jäger
2018).

We take the same approach here in Figure 5, selecting four specific
CAMEO event codings and checking how often they reflect a true
real-world event from the Crimea-Donbas synthetic narrative. We choose
four event types around key moments in the crisis. The start of the
crisis revolves around Ukraine backing out of a trade deal with the EU
in favor of Russia, but ``sign formal agreement'' events act more like a
topic detector with dozens of events generated by discussions of a
possible agreement but not the actual agreement which never
materialized. The switch is caught by the ``reject plan, agreement to
settle dispute'', but also continues for Victor Yanukovych even after he
was removed from power because of articles retroactively discussing the
cause of his removal. Events for ``use conventional military force''
capture a threshold around the start of hostilities and who the
participants were but not any particular battles or campaigns. Likewise,
``impose embargo, boycott, or sanctions'' captures the start of waves of
sanctions and from who but are effectively constant as the news coverage
does not distinguish between subtle changes or additions. In sum,
dictionary-based methods on news corpora tend to have high recall
because they parse everything in the news, but for the same reason,
their specificity for most event types is too low to back out individual
chess-like sequencing that ICBe aims to record.

@@

\begin{figure}[H]
\caption{ICBe event codings in comparison to Semantic Embeddings from source sentences\label{fig:semantic_embeddings}}
\centering{\includegraphics[width=17cm]{./p_semantic_embeddings_dendro.png}}
\textit{Notes: Dots represent individual ICB narrative sentences, as embedded by the Paraphrase-MPNET-base-v2 large language model and flattened into two dimensions with UMAP. Text labels reflect individual leaves of the ICBe ontology, and colors represent intermediate branches of the ontology. Label placement is the median of all of the sentences that the tag was applied to by the coders. The dendrogram shows hierarchical clustering of the tags. If ICBe precision is high, then the sentences that tags were applied to ought to say similar things and the intended shape of the ontology ought to be visually recognizable. }
\end{figure}
\clearpage
